# roadoi - Use oaDOI.org with R

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```


[![Build Status](https://travis-ci.org/njahn82/roadoi.svg?branch=master)](https://travis-ci.org/njahn82/roadoi)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/njahn82/roadoi?branch=master&svg=true)](https://ci.appveyor.com/project/njahn82/roadoi)
[![codecov.io](https://codecov.io/github/njahn82/roadoi/coverage.svg?branch=master)](https://codecov.io/github/njahn82/roadoi?branch=master)

roadoi interacts with the [oaDOI service](http://oadoi.org/), which links DOIs 
and open access versions of scholarly works.

API Documentation: <http://oadoi.org/api>

## How do I use it? 

Use the `oadoi_fetch()` function in this package to get open access status
information and full-text links from oaDOI. Please note that oaDOI restricts
usage to 10k requests per day. If you need more, simply contact the 
[oaDOI team](https://oadoi.org/team). They will hook you up.

```{r}
roadoi::oadoi_fetch(dois = c("10.1038/ng.3260", "10.1093/nar/gkr1047"))
```

## How do I get it? 

To install the development version, use the [devtools package](https://cran.r-project.org/web/packages/devtools/index.html)

```{r eval = FALSE}
devtools::install_github("njahn82/roadoi")
library(roadoi)
```

## Long-Form Documentation including use-case

```{r child="vignettes/intro.Rmd"}
```

## Meta

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

License: MIT

Please use the [issue tracker](https://github.com/njahn82/roadoi/issues) for bug reporting and feature requests.

