# roadoi - Use oaDOI.org with R

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```


[![Build Status](https://travis-ci.org/njahn82/roadoi.svg?branch=master)](https://travis-ci.org/njahn82/roadoi)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/njahn82/roadoi?branch=master&svg=true)](https://ci.appveyor.com/project/njahn82/roadoi)
[![codecov.io](https://codecov.io/github/njahn82/roadoi/coverage.svg?branch=master)](https://codecov.io/github/njahn82/roadoi?branch=master)
[![cran version](http://www.r-pkg.org/badges/version/roadoi)](https://cran.r-project.org/package=roadoi)
[![rstudio mirror downloads](http://cranlogs.r-pkg.org/badges/roadoi)](https://github.com/metacran/cranlogs.app)

roadoi interacts with the [oaDOI API](http://oadoi.org/), a simple interface which links DOIs 
and open access versions of scholarly works. oaDOI powers [unpaywall](http://unpaywall.org/).

API Documentation: <http://oadoi.org/api>

## How do I use it? 

Use the `oadoi_fetch()` function in this package to get open access status
information and full-text links from oaDOI.

```{r}
roadoi::oadoi_fetch(dois = c("10.1038/ng.3260", "10.1093/nar/gkr1047"))
```

There are no API restrictions. However, a rate limit of 100k is implemented. If you need to access more data, use the data dump <https://oadoi.org/api#dataset> instead.

## How do I get it? 

Install and load from [CRAN](https://cran.r-project.org/package=roadoi):

```{r eval=FALSE}
install.packages("roadoi")
library(roadoi)
```

To install the development version, use the [devtools package](https://cran.r-project.org/package=devtools)

```{r eval = FALSE}
devtools::install_github("njahn82/roadoi")
library(roadoi)
```

## Long-Form Documentation including use-case

```{r child="vignettes/intro.Rmd"}
```

## Meta

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

License: MIT

Please use the [issue tracker](https://github.com/njahn82/roadoi/issues) for bug reporting and feature requests.

