# roadoi - Use oaDOI.org with R

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```


[![Build Status](https://travis-ci.org/njahn82/roadoi.svg?branch=master)](https://travis-ci.org/njahn82/roadoi)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/njahn82/roadoi?branch=master&svg=true)](https://ci.appveyor.com/project/njahn82/roadoi)
[![codecov.io](https://codecov.io/github/njahn82/roadoi/coverage.svg?branch=master)](https://codecov.io/github/njahn82/roadoi?branch=master)

roadoi interacts with the [oaDOI service](http://oadoi.org/), which links DOIs 
and open access versions of scholarly works.

API Documentation: <http://oadoi.org/api>

## Installation

To install the development version, use the [devtools package](https://cran.r-project.org/web/packages/devtools/index.html)

```r
devtools::install_github("njahn82/roadoi")
library(roadoi)
```

## Usage

Use the `oadoi_fetch()` function in this package to get open access status
information and full-text links from oaDOI. Please note that oaDOI restricts
usage to 10k requests per day.

```{r}
roadoi::oadoi_fetch(dois = c("10.1038/ng.3260", "10.1093/nar/gkr1047"))
```

## Share your email address to support oaDOI!

oaDOI is provided by the non-profit [Impactstory](https://impactstory.org/). 
Sharing your email address with oaDOI does not only keep you updated in case 
something breaks, but also helps Impactstory reporting API usage to the non-profit's funders.

```{r}
roadoi::oadoi_fetch(dois = "10.1371/journal.pone.0018657", email = "najko.jahn@gmail.com")
```


## Where to get DOIs?

[rOpenSci](https://ropensci.org/) offers [various packages to access literature databases](https://ropensci.org/packages/#literature), and many of these packages return DOIs for scholarly works, if available.

A great tool to get a random sample set of DOIs is the 
[rcrossref](https://github.com/ropensci/rcrossref) package.

For instance, to get a random sample of 50 publications that are indexed by Crossref, a DOI-minting agency, and to explore how many of them are provided as green or gold open access:

```{r}
library(dplyr)
dois_r <- rcrossref::cr_r(sample = 50)
roadoi::oadoi_fetch(dois_r) %>% 
  group_by(oa_color) %>%
  summarise(Articles = n()) %>%
  mutate(Proportion = Articles / sum(Articles)) %>%
  knitr::kable()
```


## Meta

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

License: MIT

Please use the [issue tracker](https://github.com/njahn82/roadoi/issues) for bug reporting and feature requests.


