---
title: "Introduction to roadoi"
author: "Najko Jahn"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to roadoi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Open access copies of scholarly publications are sometimes hard to find. Some are published in open access journals. Others are made freely available as preprints before publication, and others are deposited in institutional repositories, digital archives maintained by universities and research institutions. This document guides you to roadoi, a R client that makes it easy to search for these open access copies by interfacing the [Unpaywall](https://unpaywall.org/) service where DOIs are matched with freely available full-texts available from open access journals and archives.

### About Unpaywall

[Unpaywall](https://unpaywall.org/), developed and maintained by the [team of Impactstory](https://profiles.impactstory.org/about), is a non-profit service that finds open access copies of scholarly literature simply by looking up a DOI (Digital Object Identifier). It not only returns open access full-text links, but also helpful metadata about the open access status of a publication such as licensing or provenance information.

Unpaywall uses different data sources to find open access full-texts including:

- [Crossref](http://www.crossref.org/): a DOI registration agency serving major scholarly publishers.
- [Directory of Open Access Journals (DOAJ)](https://doaj.org/): a registry of open access journals
- Various OAI-PMH metadata sources. OAI-PMH is a protocol often used by open access journals and repositories such as arXiv and PubMed Central.

See [Piwowar et al. (2018)](https://doi.org/10.7717/peerj.4375) for a comprehensive overview of Unpaywall.

### Basic usage

There is one major function to talk with Unpaywall, `oadoi_fetch()`, taking a character vector of DOIs and your email address as required arguments.

```{r}
library(roadoi)
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com")
```

#### What's returned?

The client supports API version 2. According to the [Unpaywall Data Format](http://unpaywall.org/data-format), the following variables with the following definitions are returned:

**Column**|**Description**
|:------------|:----------------------------------------------
`doi`|DOI (always in lowercase)
`best_oa_location`|list-column describing the best OA location. Algorithm prioritizes publisher hosted content (e.g. Hybrid or Gold)
`oa_locations`|list-column of all the OA locations. 
`data_standard`|Indicates the data collection approaches used for this resource. `1` mostly uses Crossref for hybrid detection. `2` uses more comprehensive hybrid detection methods. 
`is_oa`|Is there an OA copy (logical)? 
`genre`|Publication type
`oa_status`|Classifies OA resources by location and license terms as one of: gold, hybrid, bronze, green or closed. See here for more information <https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean->.
`has_repository_copy`|Is a full-text available in a repository?
`journal_is_oa`|Is the article published in a fully OA journal? Uses the Directory of Open Access Journals (DOAJ) as source. 
`journal_is_in_doaj`|Is the journal listed in the Directory of Open Access Journals (DOAJ).
`journal_issns`|ISSNs, i.e. unique code to identify journals.
`journal_issns_l`|Linking ISSN.
`journal_name`|Journal title
`publisher`|Publisher
`title`|Publication title. 
`year`|Year published. 
`published_date`|Date published.
`updated`|Time when the data for this resource was last updated. 
`authors`|Lists authors (if available)

The columns  `best_oa_location` and  `oa_locations` are list-columns
that contain useful metadata about the OA sources found by Unpaywall These are

**Column**|**Description**
|:------------|:----------------------------------------------
`evidence`|How the OA location was found and is characterized by Unpaywall?
`host_type`|OA full-text provided by `publisher` or `repository`. 
`license`|The license under which this copy is published
`url`|The URL where you can find this OA copy.
`versions`|The content version accessible at this location following the DRIVER 2.0 Guidelines  (<https://wiki.surfnet.nl/display/DRIVERguidelines/DRIVER-VERSION+Mappings>)

Note that Unpaywall schema is only informally described. Check also with <https://unpaywall.org/data-format>.

There at least [two ways to simplify these list-columns](http://r4ds.had.co.nz/many-models.html#simplifying-list-columns).

To get the full-text links from the list-column `best_oa_location`, you may want to use `purrr::map_chr()`.

```{r}
library(dplyr)
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com") %>%
  dplyr::mutate(
    urls = purrr::map(best_oa_location, "url") %>% 
                  purrr::map_if(purrr::is_empty, ~ NA_character_) %>% 
                  purrr::flatten_chr()
                ) %>%
  .$urls
```

If you want to gather all full-text links and to explore where these links are hosted, simplify the list-column `oa_locations` with `tidyr::unnest()`. Note the  column `updated`, which belongs to the main data.frame and the nested list-column. It will cause an error when flatting into regular columns. Either de-select `updated` or change the argument `names_repair`.

```{r}
library(dplyr)
library(tidyr)
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com") %>%
  tidyr::unnest(oa_locations, names_repair = "universal") %>% 
  dplyr::mutate(
    hostname = purrr::map(url, httr::parse_url) %>% 
                  purrr::map_chr(., "hostname", .null = NA_integer_)
                ) %>% 
  dplyr::mutate(hostname = gsub("www.", "", hostname)) %>%
  dplyr::group_by(hostname) %>%
  dplyr::summarize(hosts = n())
```


Note that fields to be returned might change according to the [Unpaywall API specs](http://unpaywall.org/products/api)

#### Any API restrictions?

There are no API restrictions. However, providing your email address when using this client is required by Unpaywall. Set email address in your `.Renviron` file with
the option `roadoi_email` 

```
roadoi_email = "najko.jahn@gmail.com"
```.

You can open your `.Renviron` file calling 

```r
file.edit("~/.Renviron")`
```

Save the file and restart your R session. To stop sharing your email when using rcrossref, delete it from your `.Renviron` file.

#### Keeping track of crawling

To follow your API call, and to estimate the time until completion, use the `.progress` parameter inherited from `plyr` to display a progress bar.

```{r}
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com", 
                    .progress = "text")
```

#### Catching errors

Unpaywall is a reliable API. However, this client follows [Hadley Wickham's Best practices for writing an API package](https://CRAN.R-project.org/package=httr/vignettes/api-packages.html) and throws an error when the API does not return valid JSON or is not available. To catch these errors, you may want to use [purrr's `safely()`](https://purrr.tidyverse.org/reference/safely.html) function

```{r}
random_dois <-  c("ldld", "10.1038/ng.3260", "§dldl  ")
my_data <- purrr::map(random_dois, 
              .f = purrr::safely(function(x) roadoi::oadoi_fetch(x, email = "najko.jahn@gmail.com")))
# return results as data.frame
purrr::map_df(my_data, "result")
#show errors
purrr::map(my_data, "error")

```

### Use Case: Studying the compliance with open access policies

An increasing number of universities, research organisations and funders have launched open access policies in recent years. Using roadoi together with other R-packages makes it easy to examine how and to what extent researchers comply with these policies in a reproducible and transparent manner. In particular, the [rcrossref package](https://github.com/ropensci/rcrossref), maintained by rOpenSci, provides many helpful functions for this task.

#### Gathering DOIs representing scholarly publications

DOIs have become essential for referencing scholarly publications, and thus many digital libraries and institutional databases keep track of these persistent identifiers. For the sake of this vignette, instead of starting with a pre-defined set of publications originating from these sources, we simply generate a random sample of 50 DOIs registered with Crossref by using the [rcrossref package](https://github.com/ropensci/rcrossref).

```{r, message=FALSE}
library(dplyr)
library(rcrossref)
# get a random sample of DOIs and metadata describing these works
random_dois <- rcrossref::cr_r(sample = 50) %>%
  rcrossref::cr_works() %>%
  .$data
random_dois
```

Let's see when these random publications were published

```{r}
random_dois %>%
  # convert to years
  mutate(issued = lubridate::parse_date_time(issued, c('y', 'ymd', 'ym'))) %>%
  mutate(issued = lubridate::year(issued)) %>%
  group_by(issued) %>%
  summarize(pubs = n()) %>%
  arrange(desc(pubs))
```

and of what type they are

```{r}
random_dois %>%
  group_by(type) %>%
  summarize(pubs = n()) %>%
  arrange(desc(pubs))
```

#### Calling Unpaywall

Now let's call Unpaywall. We are capturing possible errors.

```{r}
oa_df <- purrr::map(random_dois$doi, .f = purrr::safely(
  function(x) roadoi::oadoi_fetch(x, email = "najko.jahn@gmail.com")
  )) %>%
  purrr::map_df("result")
```

and merge the resulting information about open access full-text links with parts of our Crossref metadata-set

```{r}
my_df <- random_dois %>%
  select(doi, type) %>% 
  left_join(oa_df, by = c("doi" = "doi"))
```

#### Reporting

After gathering the data, reporting with R is very straightforward. You can even generate dynamic reports using [R Markdown](http://rmarkdown.rstudio.com/) and related packages, thus making your study reproducible and transparent for others.

To display how many full-text links were found and which sources were used in a nicely formatted markdown-table using the [`knitr`](https://yihui.name/knitr/)-package:

```{r, results='asis'}
my_df %>%
  group_by(is_oa) %>%
  summarise(Articles = n()) %>%
  mutate(Proportion = Articles / sum(Articles)) %>%
  arrange(desc(Articles)) %>%
  knitr::kable()
```

How did Unpaywall find those Open Access full-texts, which were characterized as best matches, and how are these OA types distributed over publication types?

```{r, results='asis'}
my_df %>%
  filter(is_oa == TRUE) %>%
  select(best_oa_location, type) %>%
  tidyr::unnest(best_oa_location) %>% 
  group_by(evidence, type) %>%
  summarise(Articles = n()) %>%
  arrange(desc(Articles)) %>%
  knitr::kable()
```

#### More examples

For more  examples, see Piwowar et al. 2018. Together with the article, they shared their analysis of Unpaywall Data as [R Markdown supplement](https://github.com/Impactstory/oadoi-paper1/).

### References 

Piwowar, H., Priem, J., Larivière, V., Alperin, J. P., Matthias, L., Norlander, B., … Haustein, S. (2018). The state of OA: a large-scale analysis of the prevalence and impact of Open Access articles. PeerJ, 6, e4375. <https://doi.org/10.7717/peerj.4375>
