---
title: "Introduction"
author: "Najko Jahn"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### About Unpaywall

[Unpaywall](https://unpaywall.org/), developed and maintained by the [team of OurResearch](https://ourresearch.org/team/about), is a non-profit service that finds open access copies of scholarly literature by looking up a DOI (Digital Object Identifier). It not only returns open access full-text links, but also helpful metadata about the open access status of a publication such as licensing or provenance information.

Unpaywall uses different data sources to find open access full-texts including:

- [Crossref](https://www.crossref.org/): a DOI registration agency serving major scholarly publishers.
- [Directory of Open Access Journals (DOAJ)](https://doaj.org/): a registry of open access journals
- Various OAI-PMH metadata sources. OAI-PMH is a protocol often used by open access journals and repositories such as arXiv and PubMed Central.

See [Piwowar et al. (2018)](https://doi.org/10.7717/peerj.4375) for a comprehensive overview of Unpaywall.

### Basic usage

There is one major function to talk with Unpaywall, `oadoi_fetch()`, taking a character vector of DOIs and your email address as required arguments.

```{r}
library(roadoi)
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com")
```

#### What's returned?

The client supports API version 2. According to the [Unpaywall Data Format](https://unpaywall.org/data-format), the following variables with the following definitions are returned:

**Column**|**Description**
|:------------|:----------------------------------------------
`doi`|DOI (always in lowercase)
`best_oa_location`|list-column describing the best OA location. Algorithm prioritizes publisher hosted content (e.g. Hybrid or Gold)
`oa_locations`|list-column of all the OA locations. 
`oa_locations_embargoed` | list-column of locations expected to be available in the future based on information like license metadata and journals' delayed OA policies
`data_standard`|Indicates the data collection approaches used for this resource. `1` mostly uses Crossref for hybrid detection. `2` uses more comprehensive hybrid detection methods. 
`is_oa`|Is there an OA copy (logical)? 
`is_paratext`| Is the item an ancillary part of a journal, like a table of contents? See here for more information <https://support.unpaywall.org/support/solutions/articles/44001894783>. 
`genre`|Publication type
`oa_status`|Classifies OA resources by location and license terms as one of: gold, hybrid, bronze, green or closed. See here for more information <https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean->.
`has_repository_copy`|Is a full-text available in a repository?
`journal_is_oa`|Is the article published in a fully OA journal? Uses the Directory of Open Access Journals (DOAJ) as source. 
`journal_is_in_doaj`|Is the journal listed in the Directory of Open Access Journals (DOAJ).
`journal_issns`|ISSNs, i.e. unique code to identify journals.
`journal_issn_l`|Linking ISSN.
`journal_name`|Journal title
`publisher`|Publisher
`published_date`|Date published
`year`|Year published. 
`title`|Publication title. 
`updated_resource`|Time when the data for this resource was last updated. 
`authors`|Lists authors (if available)

The columns  `best_oa_location` and  `oa_locations` are list-columns that contain useful metadata about the OA sources found by Unpaywall These are

**Column**|**Description**
|:------------|:----------------------------------------------
`endpoint_id`|Unique repository identifier
`evidence`|How the OA location was found and is characterized by Unpaywall?
`host_type`|OA full-text provided by `publisher` or `repository`. 
`is_best`|Is this location the \code{best_oa_location} for its resource?
`license`|The license under which this copy is published
`oa_date`|When this document first became available at this location
`pmh_id`|OAI-PMH endpoint where we found this location
`repository_institution`|Hosting institution of the repository.
`updated`|Time when the data for this location was last updated
`url`|The URL where you can find this OA copy.
`url_for_landing_page`| The URL for a landing page describing this OA copy.
`url_for_pdf`|The URL with a PDF version of this OA copy.
`versions`|The content version accessible at this location following the DRIVER 2.0 Guidelines  (<https://wiki.surfnet.nl/display/DRIVERguidelines/DRIVER-VERSION+Mappings>)

The Unpaywall schema is also described here: <https://unpaywall.org/data-format>.

The columns  `best_oa_location`. `oa_locations` and `oa_locations_embargoed` are list-columns that contain useful metadata about the OA sources found by Unpaywall. 

If `.flatten = TRUE` the list-column `oa_locations` will be restructured in a long format where each OA fulltext is represented by one row, which allows to take into account all OA locations found by Unpaywall in a data analysis.

```{r}
library(dplyr)
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814",
                             "10.1093/reseval/rvaa038",
                             "10.1101/2020.05.22.111294",
                             "10.1093/bioinformatics/btw541"), 
                    email = "najko.jahn@gmail.com", .flatten = TRUE) %>%
  dplyr::count(is_oa, evidence, is_best) 
```


#### Any API restrictions?

There are no API restrictions. However, Unpaywall requires an email address when using its API. If you are too tired to type in your email address every time, you can store the email  in the `.Renviron` file with the option `roadoi_email` 

```
roadoi_email = "najko.jahn@gmail.com"
```

You can open your `.Renviron` file calling 

```r
file.edit("~/.Renviron")`
```

Save the file and restart your R session. To stop sharing the email when using roadoi, delete it from your `.Renviron` file.

#### Keeping track of crawling

To follow your API call, and to estimate the time until completion, use the `.progress` parameter inherited from `plyr` to display a progress bar.

```{r}
roadoi::oadoi_fetch(dois = c("10.1186/s12864-016-2566-9",
                             "10.1103/physreve.88.012814"), 
                    email = "najko.jahn@gmail.com", 
                    .progress = "text")
```


### References 

Piwowar, H., Priem, J., Larivière, V., Alperin, J. P., Matthias, L., Norlander, B., … Haustein, S. (2018). The state of OA: a large-scale analysis of the prevalence and impact of Open Access articles. PeerJ, 6, e4375. <https://doi.org/10.7717/peerj.4375>
